---
title: "Running the Yeast Prompt-Ladder Evaluation"
author:
  - name: Claude code
package: BiocAgentSkills
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Running the Yeast Prompt-Ladder Evaluation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      message = FALSE, warning = FALSE)
```

# Overview

This vignette walks through the complete skill evaluation workflow for the
`load-gene-from-gff-gtf` skill using *S. cerevisiae* Ensembl r113 GFF3 as
ground truth.  We send each prompt in the P1–P6 ladder to Claude (via the
`claude` CLI), collect the generated R code, execute it against the real GFF3,
and assemble a performance table.

The workflow has three layers:

1. **Prompt layer** — six prompts ranging from minimal to adversarial
2. **Execution layer** — the `claude` CLI in non-interactive (`-p`) mode
   generates code; `eval_code_string_yeast()` evaluates it
3. **Assessment layer** — static (code-text) and dynamic (TxDb correctness)
   checks are combined into a summary table

**Authentication**: this vignette calls the `claude` CLI, which uses your
existing Claude Code authentication (no `ANTHROPIC_API_KEY` env var required).
Results are saved to `inst/eval1/eval/prompt_ladder_cache.rds`; subsequent
knits load from cache without touching the API.

```{r prereq-check}
has_cli <- nchar(Sys.which("claude")) > 0

cache_path <- file.path(
  system.file("eval1", "eval", package = "BiocAgentSkills"),
  "prompt_ladder_cache.rds"
)
has_cache <- file.exists(cache_path)
can_run   <- has_cli || has_cache

if (!can_run)
  message("claude CLI not found and no pre-built cache. ",
          "Install Claude Code and re-knit to run live.")
if (!has_cli && has_cache)
  message("Using pre-built results cache.")
```

# Setup

## Source the evaluator

`run_eval_yeast.R` auto-detects its own directory to locate sibling files.
That detection relies on `sys.frames()[[1]]$ofile`, which is not populated
inside a knitr chunk.  We work around this by temporarily setting the working
directory to the parent folder so the script's relative-path fallback
(`"eval/ground_truth_yeast.R"`) resolves correctly.

```{r source-eval}
eval_dir <- system.file("eval1", "eval", package = "BiocAgentSkills")
old_wd <- setwd(dirname(eval_dir))          # parent = eval1/
source(file.path(eval_dir, "run_eval_yeast.R"))
setwd(old_wd)
```

## Resolve the bundled yeast GFF3

```{r gff-path}
gff_path <- file.path(eval_dir, "data",
                      "Saccharomyces_cerevisiae.R64-1-1.113.gff3.gz")
stopifnot(file.exists(gff_path))
cat("GFF3:", gff_path, "\n")
cat("Size:", round(file.size(gff_path) / 1e6, 1), "MB\n")
```

## Load the skill document (P4 system prompt)

```{r skill-doc}
skill_path <- file.path(
  system.file("eval1", "how-tos", package = "BiocAgentSkills"),
  "load-gene-from-gff-gtf.md"
)
skill_doc <- paste(readLines(skill_path), collapse = "\n")
cat(substr(skill_doc, 1, 250), "...\n")
```

# The Prompt Ladder

We build the six prompts as a named list, substituting the real GFF3 path for
`<LOCAL_PATH>` and injecting the skill document into the P4 system prompt.

```{r define-prompts}
make_prompts <- function(gff, skill) {
  list(

    P1_minimal = list(
      system = NULL,
      user   = sprintf(
        "Load a GFF file for yeast in R and show me the genes and exons.\n%s",
        sprintf("Use this file: %s", gff))
    ),

    P2_pkg_hint = list(
      system = NULL,
      user   = sprintf(
        "Use txdbmaker to load this S. cerevisiae GFF3 file in R\nand extract exons grouped by gene:\n  %s",
        gff)
    ),

    P3_skill_aligned = list(
      system = NULL,
      user   = sprintf(paste0(
        "Using txdbmaker's makeTxDbFromGFF, load this S. cerevisiae GFF3 file:\n  %s\n\n",
        "Set organism=\"Saccharomyces cerevisiae\" and taxonomyId=4932.\n",
        "Show me:\n1. Exons grouped by gene\n2. Transcripts grouped by gene\n",
        "3. CDS grouped by transcript"), gff)
    ),

    P4_full_skill = list(
      system = sprintf(paste0(
        "You are an expert Bioconductor developer. ",
        "Use the following how-to guide when answering questions ",
        "about loading gene models:\n\n---\n%s\n---"), skill),
      user   = sprintf(paste0(
        "Load this S. cerevisiae GFF3 file using Bioconductor ",
        "and show exons, transcripts, and CDS grouped appropriately.\n",
        "Set organism=\"Saccharomyces cerevisiae\" and taxonomyId=4932.\n",
        "File: %s"), gff)
    ),

    P5_remote_url = list(
      system = NULL,
      user   = paste0(
        "Use txdbmaker to load the S. cerevisiae gene annotation directly from ",
        "Ensembl release 113 and show exons grouped by gene:\n  ",
        "https://ftp.ensembl.org/pub/release-113/gff3/saccharomyces_cerevisiae/",
        "Saccharomyces_cerevisiae.R64-1-1.113.gff3.gz")
    ),

    P6_adversarial = list(
      system = NULL,
      user   = sprintf(
        "Use GenomicFeatures to load this yeast GFF3 file as a TxDb, then show exons grouped by gene:\n  %s",
        gff)
    )
  )
}

prompts <- make_prompts(gff_path, skill_doc)
cat("Prompts:", paste(names(prompts), collapse = ", "), "\n")
```

# Generating Responses via the Claude CLI

## Helper functions

`ask_claude()` shells out to `claude -p` in non-interactive mode.  All tools
are disabled (`--tools ""`) so the model returns only text; it cannot execute
code or read files.  `extract_r_code()` pulls the first fenced R code block
from the markdown response.

```{r helpers}
ask_claude <- function(user_msg, system_msg = NULL, model = "haiku") {
  claude <- Sys.which("claude")
  if (nchar(claude) == 0L) stop("claude CLI not found on PATH")

  args <- c("-p", "--model", model, "--tools", "")
  if (!is.null(system_msg))
    args <- c(args, "--system-prompt", system_msg)

  # Pass the user message via stdin (the `input` argument to system2)
  out <- system2(claude, args = args,
                 input  = user_msg,
                 stdout = TRUE, stderr = FALSE)
  paste(out, collapse = "\n")
}

extract_r_code <- function(text) {
  m <- regmatches(text,
        regexpr("```[rR]\n(.*?)```", text, perl = TRUE))
  if (length(m) == 0L || nchar(m) == 0L) return(text)
  sub("^```[rR]\n", "", sub("```$", "", m))
}
```

## Running the prompt ladder

For each of the six prompts we: call the Claude CLI to get generated code,
extract the R code block, run `eval_code_string_yeast()`, and record both the
generated code and the full result object.

`P5_remote_url` asks Claude to load the GFF3 directly from the Ensembl FTP
server; if network access is restricted the generated code will fail in the
dynamic check while static checks still run.

```{r run-ladder, eval=has_cli && !has_cache}
run_one <- function(prompt_name, prompt, model = "haiku") {
  message("  ", prompt_name, " ... ", appendLF = FALSE)
  t0 <- proc.time()

  response <- tryCatch(
    ask_claude(prompt$user, prompt$system, model = model),
    error = function(e) paste("CLI ERROR:", conditionMessage(e))
  )
  code   <- extract_r_code(response)
  result <- eval_code_string_yeast(code, verbose = FALSE)

  elapsed <- round((proc.time() - t0)[["elapsed"]], 1)
  message(if (result$pass) "PASS" else "FAIL", " (", elapsed, "s)")

  list(prompt_name = prompt_name,
       response    = response,
       code        = code,
       result      = result,
       elapsed_s   = elapsed)
}

message("Running prompt ladder (model = haiku) ...")
ladder_results <- lapply(names(prompts),
                         function(nm) run_one(nm, prompts[[nm]]))
names(ladder_results) <- names(prompts)

saveRDS(ladder_results, cache_path)
message("Saved to: ", cache_path)
```

## Load cached results

```{r load-cache, eval=can_run}
if (!exists("ladder_results") && has_cache)
  ladder_results <- readRDS(cache_path)
```

# Inspecting Generated Code

Before aggregating, it is instructive to see what code each prompt level
actually produced.  The differences reveal precisely how much guidance the
model needs.

```{r show-code, eval=can_run, results='asis'}
for (nm in names(ladder_results)) {
  cat(sprintf("\n## %s\n\n", nm))
  cat("````r\n")
  cat(ladder_results[[nm]]$code)
  cat("\n````\n\n")
}
```

# Evaluation Results

## Per-prompt check tables

```{r per-prompt-tables, eval=can_run, results='asis'}
for (nm in names(ladder_results)) {
  res <- ladder_results[[nm]]$result
  cat(sprintf("\n### %s  (overall: **%s**)\n\n",
              nm, if (res$pass) "PASS" else "FAIL"))
  tbl <- res$combined
  tbl$status <- ifelse(tbl$status == "PASS", "✅ PASS", "❌ FAIL")
  print(knitr::kable(tbl[, c("check","status","detail")],
                     col.names = c("Check","Status","Detail"),
                     row.names = FALSE))
  cat("\n")
}
```

## Summary: static checks

These checks are applied to the code text before execution, so they run even
when the code itself fails.

```{r static-summary, eval=can_run}
static_keys <- c("uses txdbmaker", "avoids deprecated path", "format= arg",
                  "exonsBy", "transcriptsBy", "cdsBy")

static_mat <- sapply(ladder_results, function(r) {
  s <- r$result$static
  v <- setNames(s$status == "PASS", s$check)
  v[static_keys]
})

sdf <- as.data.frame(t(static_mat))
sdf$Prompt <- rownames(sdf)
rownames(sdf) <- NULL
sdf <- sdf[, c("Prompt", static_keys)]
names(sdf) <- c("Prompt", "txdbmaker", "no-deprecated",
                 "format=", "exonsBy", "transcriptsBy", "cdsBy")
for (col in names(sdf)[-1])
  sdf[[col]] <- ifelse(sdf[[col]], "✅", "❌")

knitr::kable(sdf, row.names = FALSE,
             caption = "Static check results by prompt (P1–P6)")
```

## Summary: correctness checks

Correctness checks execute the code and compare the resulting `TxDb` against
`YEAST_FACTS` (7 036 genes, 17 seqlevels, 5 named landmark genes).

```{r correctness-summary, eval=can_run}
cdf <- data.frame(
  Prompt    = names(ladder_results),
  N_checks  = sapply(ladder_results, function(r) nrow(r$result$correctness)),
  N_pass    = sapply(ladder_results, function(r)
                sum(r$result$correctness$status == "PASS")),
  Overall   = sapply(ladder_results, function(r) r$result$pass),
  Elapsed_s = sapply(ladder_results, function(r) r$elapsed_s),
  stringsAsFactors = FALSE
)
cdf$Pct_pass <- round(100 * cdf$N_pass / cdf$N_checks, 1)
cdf$Overall  <- ifelse(cdf$Overall, "✅ PASS", "❌ FAIL")

knitr::kable(
  cdf[, c("Prompt","N_checks","N_pass","Pct_pass","Overall","Elapsed_s")],
  col.names = c("Prompt","# Checks","# Pass","% Pass","Overall","Time (s)"),
  row.names = FALSE,
  caption   = "Correctness check results by prompt (P1–P6)"
)
```

## Prompt-level pass rate chart

```{r pass-chart, eval=can_run, fig.width=7.5, fig.height=4.5,
    fig.cap="Fraction of correctness checks passing by prompt level. Dashed line at 80%."}
pct  <- as.numeric(cdf$Pct_pass)
cols <- ifelse(pct >= 80, "#2ca02c",
        ifelse(pct >= 40, "#ff7f0e", "#d62728"))

op <- par(mar = c(7, 4, 3, 1))
bp <- barplot(pct,
  names.arg = cdf$Prompt,
  col       = cols,
  ylim      = c(0, 115),
  las       = 2,
  ylab      = "% correctness checks passing",
  main      = "Skill performance by prompt level (claude haiku)",
  cex.names = 0.8)
abline(h = 80, lty = 2, col = "grey50")
text(bp, pct + 4, sprintf("%.0f%%", pct), cex = 0.85, font = 2)
legend("topright",
       legend = c("≥80%  (good)", "40–79%  (partial)", "<40%  (failing)"),
       fill   = c("#2ca02c","#ff7f0e","#d62728"),
       bty    = "n", cex = 0.85)
par(op)
```

# Interpreting the Results

## The prompt ladder effect

The six prompts isolate distinct sources of model behaviour:

| Prompt | What it isolates |
|--------|-----------------|
| **P1** Minimal | Baseline: does the model choose `txdbmaker` without any hint? |
| **P2** Package hint | Value of naming the package: does the model use `makeTxDbFromGFF` correctly? |
| **P3** Skill-aligned | Value of the reference pattern: function name + file path + all three grouping calls |
| **P4** Full skill doc | Upper bound: model has the complete how-to in its context window |
| **P5** Remote URL | Generalization: can the model adapt the skill to a URL rather than a local path? |
| **P6** Adversarial | Robustness: does the skill override an explicit request for the deprecated `GenomicFeatures` path? |

## Key metrics

**Skill leverage** = gap between P1 and P4.  A large gap means the skill
document is delivering real value.  A small gap means the model already knows
the pattern and the skill is low-priority for this task.

**Incremental document value** = gap between P3 and P4.  If P3 ≈ P4, a
carefully-worded prompt is sufficient and the document adds little.  If
P4 >> P3, the document contains detail that no prompt alone can supply.

**Robustness** = P6 pass rate.  A fully robust model should redirect the
`GenomicFeatures` framing to `txdbmaker::makeTxDbFromGFF` even when the user
explicitly asks for the deprecated path.  Failure here reveals a gap in the
skill's ability to override user priors.

## What to do with the results

- If P1 already passes most checks: the model has memorised the pattern from
  training data; the skill may be redundant for this task.
- If P3 fails but P4 passes: the skill document contains critical detail
  (e.g., the exact `system.file()` path, the `format=` argument) that users
  do not spontaneously include in prompts.
- If P6 fails: add an explicit deprecation notice to the skill document and
  consider restructuring the opening to name `txdbmaker` before any mention of
  `GenomicFeatures`.

# Session Info

```{r session-info}
sessionInfo()
```
