---
title: "Evaluating a skill: creating gene models from GFF3"
author:
  - name: Claude code
  - name: Vince Carey
package: BiocAgentSkills
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Evaluating a skill: creating gene models from GFF3}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


# Overview

This vignette walks through a skill evaluation workflow for the
`load-gene-from-gff-gtf` skill using a limited GFF3 file that is shipped with txdbmaker.

The workflow has three layers:

1. **Prompt layer** — six prompts ranging from minimal to adversarial
2. **Execution layer** — the `claude` CLI in non-interactive (`-p`) mode
   generates code; `eval_code_string_yeast()` evaluates it
3. **Assessment layer** — static (code-text) and dynamic (TxDb correctness)
   checks are combined into a summary table

## Resolve the bundled GFF3

```{r gff-path}
gff_path <- system.file("extdata", "GFF3_FILES", "a.gff3", package="txdbmaker")
stopifnot(file.exists(gff_path))
cat("GFF3:", gff_path, "\n")
cat("Size:", round(file.size(gff_path) / 1e6, 1), "MB\n")
```

## Load the skill document

```{r skill-doc}
skill_path <- file.path(
  system.file("demoskills", "genemodel.md", package = "BiocAgentSkills")
)
skill_doc <- paste(readLines(skill_path), collapse = "\n")
cat(substr(skill_doc, 1, 250), "...\n")
```

## Generate code according to a prompt

This code is not evaluated in the vignette, but
should succeed if the environment variable ANTHROPIC_API_KEY
is validly set.
```
syspr = readLines(system.file("demoskills", 
    "genemodel.md", package="BiocAgentSkills")) |> paste(collapse="\n")

library(ellmer)
library(btw)

k = Sys.getenv("ANTHROPIC_API_KEY")
stopifnot(nchar(k)>0)

ch = chat_anthropic(system_prompt = syspr)

newprompt = readLines(system.file("evaluation_support", "prompt2.txt", package="BiocAgentSkills")) |> paste(collapse="\n")

ch$chat(btw(), newprompt)
```

## Evaluate the generated code

The following report is produced by using
materials in inst/evaluation_support, according to the
README.md there.

```
=== Exons grouped by gene ===

=== Transcripts grouped by gene ===

=== CDS grouped by transcript ===
0/0 packages newly attached/loaded, see sessionInfo() for details.

=== Static checks ===
                  check status detail
         uses txdbmaker   PASS       
 avoids deprecated path   PASS       
            format= arg   PASS       
                exonsBy   PASS       
          transcriptsBy   PASS       
                  cdsBy   PASS       

=== Correctness checks ===
               check status                               detail
             is TxDb   PASS                                     
          gene count   PASS                   488 (expected 488)
    transcript count   PASS                   488 (expected 488)
          exon count   PASS                 1268 (expected 1268)
          CDS groups   PASS                   488 (expected 488)
           seqlevels   PASS 'SL2.40ch00' (expected 'SL2.40ch00')
         gene1 start   PASS                                16437
           gene1 end   PASS                                18189
        gene1 strand   PASS                                    +
    gene1 exon count   PASS                                    2
   gene1 exon1 start   PASS                                16437
     gene1 exon1 end   PASS                                17275
   gene1 exon2 start   PASS                                17336
     gene1 exon2 end   PASS                                18189
      gene1 tx count   PASS                                    1
       gene1 tx name   PASS                   Solyc00g005000.2.1
  gene3ex exon count   PASS                       3 (expected 3)
 gene3ex exon starts   PASS                    68062,68344,68654
  gene4ex exon count   PASS                       4 (expected 4)
  gene5ex exon count   PASS                       5 (expected 5)
      gene5ex strand   PASS                                    -

Overall: PASS  (27/27 checks passed)
```
